Computing solution space properties of combinatorial optimization problems via generic tensor networks

### 2022-09-02 paper/paper.tex
update

\section{Tensor networks}\label{sec:tn}
A tensor network is composed of a collection of tensors $\mathcal{T}$, an indicator of which tensor is the output, and a map $\lambda$ from tensor to labels associated with tensor dimensions.
\begin{definition}[Tensor Network~\cite{Cirac2021, Orus2014}]
    A tensor network is a quadruple of $\mathcal{N} = (\Tensors, \Sigma, O \in \Tensors, \Inputs : \mathcal{T} \rightarrow \Sigma^k)$. $\Tensors$ is a set of tensors, among which $O$ is the output tensor.
    $\Sigma$ is an alphabet of labels. $\Inputs$ is a map from a tensor to a vector of labels with the vector length the same as the tensor rank.
\end{definition}
The operation to evaluate a tensor network is called \textit{contraction}, which is defined as a summation of tensor element products over the labels not appearing in the output tensor.
For example, the matrix multiplication is a special tensor network
\begin{equation}
\mathcal{N}_{\rm matmul} = \left(\{A, B, C\}, \{i,j,k\}, C, T \mapsto \begin{cases} (i,j), T = A\\(j, k), T=B\\ (i, k), T=C\end{cases}\right),
\end{equation}
where $A, B$ and $C$ are matrices (two-dimensional tensors), among which $C$ is the output tensor, $(i,k),(i,j)$ and $(i,k)$ are labels associated with them. The contraction is defined as $C_{ik} = \sum_j A_{ij}B_{jk}$, where the subscripts are for tensor indexing, and the tensor dimensions with a same label must have the same size.
%can be viewed as a generalization of binary matrix multiplication to nary tensor contraction.
%Generalized Einstein's notation is often used to represent a tensor network, where we use a label in the subscript to represent a degree of freedom.
%One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
% Einstein's notation is often used to represent a tensor network, e.g.\ it represents the matrix multiplication between two matrices $A$ and $B$ as $C_{ik} = A_{ij}B_{jk}$,
% where we use a label in the subscript to represent a degree of freedom.
% One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
% In the standard Einstein's notation for tensor networks in physics, each index appears precisely twice, either both in input tensors (which will be summed over) or one in a input tensor and another in the output tensor.
% Hence a tensor network can be represented as a open simple graph,
% where a tensor is mapped to a vertex, a label shared by two input tensors is mapped to a normal edge and a label appears in the output tensor is mapped to an open edge.
% In this work, we adopt a more general form of tensor networks that do not restrict how many times a label can appear in tensors. 
The graphical representation of a tensor network is an open (meaning an edge can connect to external vertices) hypergraph, where an input tensor is mapped to a vertex and a label is mapped to a hyperedge that can connect an arbitrary number of vertices, while the labels appearing in the output tensor are open.
Our notation is a minor generalization of the standard tensor network notation used in physics as we do not restrict the number of times a label can appear in the tensors. 
While this generalized form is equivalent in representation power, it can have smaller contraction complexity as will be illustrated in \Cref{sec:tensorbad}.
% because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors, a high dimensional equivalence of the identity matrix.
%However, introducing $\delta$ tensors may significantly increase the complexity of the tensor contraction. We illustrate this subtle point in \Cref{sec:tensorbad}.

\begin{example}
\blue{maybe remove this example?}
\begin{equation}
\begin{split}
    \mathcal{T} &= \{A, B, V, C\},\\
    \Sigma &= \{i,j,k,l,m\},\\
    O &= C,\\
    \lambda &= T \mapsto \begin{cases}
    (j, k, m), &T = A\\
    (m, i, l), &T = B \\
    (j, m), &T = V \\
    (i, j, k), &T= C
    \end{cases}
    \end{split}
\end{equation}
is a tensor network that can be evaluated as $C_{ijk} = \sum_{ml}A_{jkm} B_{mil} V_{jm}$.
Its hypergraph representation is shown below, where we use different colors to represent different hyperedges.

@@

\section{Tensor network representation of independent sets} \label{sec:tnmap}
This section describes the reduction of the independent set problem to a tensor network contraction problem.
An alternative interpretation, perhaps more accessible to physicists, can be found in \Cref{sec:energymodel}, where we introduce the reduction from the energy model of hardcore lattice gases~\cite{Dyre2016, Fernandes2007}.
Let $G=(V,E)$ be an undirected graph with each vertex $v\in V$ being associated with a weight $w_v$.
An independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, we have $(u, v) \not\in E$.
To the independent set problem on $G$ can be encoded as a tensor network $\mathcal{N}_{\rm IS}(G)$
\begin{equation}\label{eq:mistensornetwork}
\begin{split}
    \mathcal{T} &= \{W^{(v)} | v\in V\} \cup \{B^{(u, v)} | (u, v) \in E\} \cup \{P\},\\
    \Sigma &= \{s_v | v \in V\},\\
    O &= P,\\
    \lambda &= T \mapsto \begin{cases}
                    (s_v), &T = W^{(v)}\\
                    (s_u, s_v), &T = B^{(u, v)}\\
                    (), & T = P
                    \end{cases},
    \end{split}
\end{equation}
where for each vertex $v$, we define a parameterized rank-one tensor associated with it as
\begin{equation}
    W^{(v)} = \left(\begin{matrix}
        1 \\
        x_v^{w_v}
    \end{matrix}\right),\label{eq:vertextensor}
\end{equation}
and for each edge $(u, v) \in E$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B^{(u, v)} = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right). \label{eq:edgetensor}
\end{equation}
We map each vertex $v\in V$ to a label $s_v \in \{0, 1\}$ of dimension $2$, where we use $0$ ($1$) to denote a vertex is absent (present) in $I$.
%We use the superscripts to denote which vertex or edge a tensor is associated to.
These labels can be used as subscripts of tensors to index tensor elements, e.g.\ $W^{(v)}_0=1$ is the first element associated with $s_v=0$ and $W^{(v)}_1=x_v^{w_v}$ is the second element associated with $s_v=1$, where $x_v$ is a variable associated with vertex $v$.
$P$ is the output tensor of rank $0$, i.e. a scalar.
The independence constraint is encoded in the edge tensors, where we use $B_{11} = 0$ to denote two vertices connected by an edge cannot be both in the independent set. 
The contraction of this tensor network is
\begin{equation}\label{eq:idp}
    P(G) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{v\in V} W^{(v)}_{s_v} \prod\limits_{(u,v) \in E} B^{(u,v)}_{s_u s_v},
\end{equation}
where the summation runs over all $2^{|V|}$ vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$. 
% It is easy to see that an edge tensor element $B_{s_us_v}$ encodes the independent set constraint. It zeros out the product whenever both $u$ and $v$ are in the independent set.
A vertex tensor element $W^{(v)}_{s_v}$ contributes a multiplicative factor $x_v^{w_v}$ whenever $v$ is in the set.

@@

For a general graph, it is computationally inefficient to evaluate \Cref{eq:idp} by directly summing up the $2^{|V|}$ products.
A better approach to evaluate a tensor network is: to find a good pair-wise tensor contraction order as a binary tree and then contract two tensors at a time by this order.
\begin{theorem}\label{thm:complexreal}
    If the addition and multiplication of tensor elements can be computed in time $O(1)$,
    the tensor network in \Cref{eq:mistensornetwork} for the independent set problem on graph $G = (V,E)$ can be contracted in time $\cc = O(|E|)2^{O({\rm tw}(G))}$.
\end{theorem}
\begin{proof}
Let us denote the line graph, a graph obtained by mapping an edge in the original graph to a vertex and connect two vertices if and only if their associated edges in the original original graph share a common vertex, of the hypergraph representation of a tensor network $\mathcal{N}$ as $L(\mathcal{N})$.
A contraction order of $\mathcal{N}$ corresponds to a tree decomposition of $L(\mathcal{N})$ and the largest intermediate tensor has a rank equal to the width of its tree decomposition~\cite{Markov2008}.
Therefore, an optimal (in terms of space complexity) contraction order corresponds to the tree decomposition of $L(\mathcal{N})$ with the smallest width (or the treewidth ${\rm tw}(L(\mathcal{N})$). The contraction complexity is $O(|\mathcal{N}|)2^{O({\rm tw}(L(\mathcal{N})))}$, where $|\mathcal{N}|$ is the number of tensors in $\mathcal{N}$.
For the independent set problem on graph $G = (V, E)$, the line graph of the hypergraph representation of the tensor network in \Cref{eq:mistensornetwork} is the same as the graph $G$ up to some isolated vertices, hence the contraction complexity of this tensor network with an optimal contraction order is $O(|E|)2^{O({\rm tw}(G))}$.
\end{proof}

@@

\begin{lemma}
    The number of the largest $k$ independent sets of a graph $G = (V, E)$ can be computed in time $O(K\log K)\cc$.
\end{lemma}
The proof is similar to proving \Cref{thm:complexpoly}, except only $k$ largest coefficients are involved in the addition and multiplication.

@@

The configuration enumeration of independent sets include, for example, the enumeration of all independent sets, the enumeration of all MISs, and the enumeration of independent sets with the largest several sizes.
Recall that in the definition of a vertex tensor in \Cref{eq:vertextensor}, variables carry labels, so that one can read out all independent sets directly from the output polynomials.
The multiplication between labeled variables is commutative while the summation of labeled variables forms a set.
Intuitively, one can use a bit string as the representation of a labeled variable and use the bit-wise or $\lor^\circ$ as the multiplication operation.
For example, in a 5-vertex graph, $x_2$ and $x_5$ can be represented as $01000$ and $00001$ respectively and their multiplication $x_2x_5$ can be represented as $01000 \lor^\circ 00001 = 01001$.
To enumerate all independent sets, we designed an algebra on sets of bit strings:
\begin{equation}
\eqname{SN}
\begin{split}
    s \oplus t &= s \cup t\\
    s \odot t &= \{\sigma \lor^\circ \tau \, | \, \sigma \in s, \tau \in t\}\\
    \mymathbb{0} &= \{\}\\
    \mymathbb{1} &= \{0^{\otimes |V|}\},
\end{split}\label{eq:set}
\end{equation}
where $s$ and $t$ are each a set of $|V|$-bit strings.
\begin{example}\label{eg:setalgebra}
    For elements being bit strings of length $5$, we have the following set algebra
\begin{equation*}
\begin{split}
    &\{00001\} \oplus \{01110, 01000\} = \{01110, 01000\} \oplus \{00001\} = \{00001,01110, 01000\}\\
    &\{00001\} \oplus \{\} = \{00001\}\\
&\\
    &\{00001\} \odot \{01110, 01000\} = \{01110, 01000\} \odot \{00001\} = \{01111, 01001\}\\
    &\{00001\} \odot \{\} = \{\}\\
    &\{00001\} \odot \{00000\} = \{00001\}.
\end{split}
\end{equation*}
\end{example}

\begin{theorem}\label{thm:complexset}
    The independent sets of graph $G = (V, E)$ be a graph can be enumerated in time $O(|I|)\cc$, where $|I|$ is the number of independent sets.
\end{theorem}
\begin{proof}
To enumerate over $I$,  we initialize variable $x_{i}$ in the vertex tensor to $x_i = \{\boldsymbol{e}_{i}\}$, where $\boldsymbol{e}_i$ is a basis bit string of size $|V|$ that has only one non-zero value at location $i$.
The vertex and edge tensors are thus
\begin{equation}
    \left(W^{\rm SN}\right)^{(v)} = \left(\begin{matrix}
        \mymathbb{1} \\
        \{\boldsymbol{e}_{i}\}
    \end{matrix}\right),   
    \qquad \qquad
        \left(B^{\rm SN}\right)^{(u, v)} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}
The contraction of this tensor network is a set with its elements being all independent set configurations. The time complexity of element-wise addition and multiplication is upper bounded by $|I||V|$, where $|V|$ comes from the linear cost of representing a $|V|$-vertex configuration. Combining with \Cref{thm:complexreal}, the overall time complexity is $O(|I||V|)\cc$.
In practise, the huge multiplicative factor $|I|$ only appear in the last step of tensor contraction.
This complexity only upper bounds the actually performance of the algorithm.
\end{proof}

@@

For example, to enumerate only the MISs, we can define a combined element type $s_{k}\infty^k$, where the coefficients follow the algebra in \Cref{eq:set} and the exponents follow the max-plus tropical algebra.
The combined operations become: 
\begin{equation}
\eqname{P1+SN}
\begin{split}
    s_x\infty^x \oplus s_y\infty^y &= \begin{cases}
        (s_x \cup s_y)\infty^{\max(x,y)}, & x = y\\
        s_y\infty^{\max(x,y)}, & x < y\\
        s_x\infty^{\max(x,y)}, & x > y
    \end{cases},\\
    s_x\infty^x \odot s_y\infty^y &= \{\sigma \lor^\circ \tau | \sigma \in s_x, \tau \in s_y\}\infty^{x+y},\\
    \mymathbb{0} &= \{\}\infty^{-\infty},\\
    \mymathbb{1} &= \{0^{\otimes |V|}\}\infty^{0}. \label{eq:countingtropicalset}
\end{split}
\end{equation}

@@

\begin{theorem}
    Let $G = (V, E)$ be a graph, one of its maximum independent set can be computed in time $O(|V|)\cc$.
\end{theorem}
The proof is similar to \Cref{thm:complexmis}, except the set algebra is replace by the one in \Cref{eq:singleconfig}, and the time complexity of element-wise addition and multiplication no longer depends on $|\mathcal{I}_\alpha|$.
The linear dependency of $|V|$ in the time complexity can be remove using the back propagation technique, at the cost of additional space.
\begin{theorem}
    Let $G = (V, E)$ be a graph, one of its maximum independent set can be computed in time $\cc$.
\end{theorem}
\begin{proof}
    Let $\alpha(G) = \max\limits_{s \in I} \sum_{v} s_v$ be the maximum independent set size. The optimal configuration can be obtained by differentiating over vertex tensors as
    \begin{equation}
    \begin{cases}
        1, &\frac{\partial \alpha(G)}{\partial s_v} = 1\\
        0, &\frac{\partial \alpha(G)}{\partial s_v} = 0
    \end{cases}.
    \end{equation}
    It can be numerically computed by differentiating over the process of the tropical tensor network contraction, where the backward rules can be found in \Cref{sec:bounding}. The time overhead of computing single optimal solution is constant comparing to only computing the contraction. Thus the computing time complexity should be the same as in \Cref{thm:complextropical}.
\end{proof}

@@

All the solution space properties and the corresponding algebra on unweighted graphs still hold for integer-weighted graphs, while for general weighted graphs, the independence polynomial is not well defined anymore.
For general weighted graphs, it is more useful to know the $k$ maximum weighted sets and their sizes.
They can be computed by the extended tropical algebra, which is a natural generalization of the max-plus tropical algebra:
\begin{equation}
\eqname{T$k$}
\begin{split}
    s \oplus t &= \texttt{largest}(s \cup t, k)\\
    s \odot t &= \texttt{largest}(\{a+b | a \in s, b\in t\}, k)\\
    \mymathbb{0} &= -\infty^{\otimes k}\\
    \mymathbb{1} &= -\infty^{\otimes k-1} \otimes 0
\end{split}\label{eq:ext-tropical}
\end{equation}
where $\texttt{largest}(s, k)$ means truncating the set $s$ by only keeping its $k$ largest values.

@@

\subsection{Matching problem}
A $k$-matching in a graph $G=(V,E)$ is a set of $k$ edges that no two of which have a vertex in common.
We map an edge $(u, v) \in E$ to a degree of freedom $\langle u, v\rangle \in \{0, 1\}$ in a tensor network, where $1$ means an edge is in the set and $0$ otherwise.
The tensor network representation for the matching problem $\mathcal{N}_{match}$ can be defined as
\begin{equation}\label{eq:matchtensornetwork}
\begin{split}
    \mathcal{T} &= \{W^{(v)} | v\in V\} \cup \{B^{(u, v)} | (u, v) \in E\} \cup \{P\},\\
    \Sigma &= \{\langle u,v\rangle | (u, v) \in E\},\\
    O &= P,\\
    \lambda &= S \mapsto \begin{cases}
                    (\langle v, N(v, 1)\rangle, \langle v, N(v,2) \rangle, \ldots, \langle v, N(v, d(v))\rangle), & S = W^{(v)}\\
                    (\langle u,v\rangle), & S = B^{(u, v)}\\
                    (), & S = P
                    \end{cases},
\end{split}
\end{equation}
where for each $v\in V$, we define a vertex tensor over its neighborhood $N(v)$ as
\begin{equation}
    W^{(v)}_{\langle v, N(v, 1)\rangle, \langle v, N(v, 2) \rangle, \ldots, \langle v, N(v,d(v))\rangle} = \begin{cases}
        1, & \sum_{i=1}^{d(v)} \langle v, N(v, i) \rangle \leq 1,\\
        0, & \text{otherwise},
    \end{cases}
\end{equation}
and for each bond $(u, v) \in E$, we define a rank one tensor as
\begin{equation}
    B^{(u, v)}_{\langle u, v\rangle} = \begin{cases}
    1, & \langle u, v \rangle = 0 \\
    x^{w_{\langle u,v \rangle}}_{\langle u, v\rangle}, & \langle u, v \rangle = 1
\end{cases}.
\end{equation}
Here, $N(v, k)$ is the $k$th vertex in $N(v)$ and $d(v) = |N(v)|$ is the degree of vertex $v$, a label $\langle v, u \rangle$ is equivalent to $\langle u,v\rangle$.
$W$ tensor specifies the constraint that a vertex cannot be shared by two edges in the edge set, and an edge tensor carries the weights.

>>>>

\section{Tensor networks}\label{sec:tn}
A tensor network is a multi-linear map composed of a collection of labelled tensors $\mathcal{T}$ and an indicator for the output tensor. It is formally defined as follows.
\begin{definition}[Tensor Network~\cite{Cirac2021, Orus2014}]
    A tensor network is a triple of $\mathcal{N} = (\Sigma, \Tensors, O)$.
    $\Sigma$ is an alphabet of labels.
    $\Tensors$ is a set of labelled tensors as the inputs, each $T_{\boldsymbol{\sigma}} \in \mathcal{T}$ is labeled by a vector of labels $\boldsymbol{\sigma} \in \Sigma^{r(T)}$, where $r(T)$ the rank of $T$.
    $O$ is a labbeled tensor as the output.
\end{definition}
The operation to evaluate a tensor network is called \textit{contraction}, which is defined as a summation of tensor element products over the labels not appearing in the output tensor.
For example, the matrix multiplication is a special tensor network
\begin{equation}
\mathcal{N}_{\rm matmul} = \left(\{i,j,k\}, \{A_{ij}, B_{jk}\}, C_{ik}\right),
\end{equation}
where $A, B$ and $C$ are matrices (two-dimensional tensors), among which $C$ is the output tensor, $(i,k),(i,j)$ and $(i,k)$ are labels associated with them. The contraction is defined as $C_{ik} = \sum_j A_{ij}B_{jk}$, where the subscripts are for tensor indexing, and the tensor dimensions with a same label must have the same size.
%can be viewed as a generalization of binary matrix multiplication to nary tensor contraction.
%Generalized Einstein's notation is often used to represent a tensor network, where we use a label in the subscript to represent a degree of freedom.
%One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
% Einstein's notation is often used to represent a tensor network, e.g.\ it represents the matrix multiplication between two matrices $A$ and $B$ as $C_{ik} = A_{ij}B_{jk}$,
% where we use a label in the subscript to represent a degree of freedom.
% One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
% In the standard Einstein's notation for tensor networks in physics, each index appears precisely twice, either both in input tensors (which will be summed over) or one in a input tensor and another in the output tensor.
% Hence a tensor network can be represented as a open simple graph,
% where a tensor is mapped to a vertex, a label shared by two input tensors is mapped to a normal edge and a label appears in the output tensor is mapped to an open edge.
% In this work, we adopt a more general form of tensor networks that do not restrict how many times a label can appear in tensors. 
The graphical representation of a tensor network is an open (meaning an edge can connect to external vertices) hypergraph, where an input tensor is mapped to a vertex and a label is mapped to a hyperedge that can connect an arbitrary number of vertices, while the labels appearing in the output tensor are open.
Our notation is a minor generalization of the standard tensor network notation used in physics as we do not restrict the number of times a label can appear in the tensors. 
While this generalized form is equivalent in representation power, it can have smaller contraction complexity as will be illustrated in \Cref{sec:tensorbad}.
% because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors, a high dimensional equivalence of the identity matrix.
%However, introducing $\delta$ tensors may significantly increase the complexity of the tensor contraction. We illustrate this subtle point in \Cref{sec:tensorbad}.

\begin{example}
\blue{maybe remove this example?}
\begin{equation}
\begin{split}
    \Sigma &= \{i,j,k,l,m\},\\
    \mathcal{T} &= \{A_{jkm}, B_{mil}, V_{jm}\},\\
    O &= C_{ijk},\\
    \end{split}
\end{equation}
is a tensor network that can be evaluated as $C_{ijk} = \sum_{ml}A_{jkm} B_{mil} V_{jm}$.
Its hypergraph representation is shown below, where we use different colors to represent different hyperedges.

@@

\section{Tensor network representation of independent sets} \label{sec:tnmap}
This section describes the reduction of the independent set problem to a tensor network contraction problem.
An alternative interpretation, perhaps more accessible to physicists, can be found in \Cref{sec:energymodel}, where we introduce the reduction from the energy model of hardcore lattice gases~\cite{Dyre2016, Fernandes2007}.
Let $G=(V,E)$ be an undirected graph with each vertex $v\in V$ being associated with a weight $w_v$.
An independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, we have $(u, v) \not\in E$.
To the independent set problem on $G$ can be encoded as a tensor network $\mathcal{N}_{\rm IS}(G)$
\begin{equation}\label{eq:mistensornetwork}
\begin{split}
    \Sigma &= \{s_v \mid v \in V\},\\
    \mathcal{T} &= \{W^{(v)}_{v} \mid v\in V\} \cup \{B^{(u, v)}_{uv} \mid (u, v) \in E\},\\
    O &= P,
    \end{split}
\end{equation}
where for each vertex $v$, we define a parameterized rank-one tensor associated with it as
\begin{equation}
    W^{(v)} = \left(\begin{matrix}
        1 \\
        x_v^{w_v}
    \end{matrix}\right),\label{eq:vertextensor}
\end{equation}
and for each edge $(u, v) \in E$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B^{(u, v)} = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right). \label{eq:edgetensor}
\end{equation}
We map each vertex $v\in V$ to a label $s_v \in \{0, 1\}$ of dimension $2$, where we use $0$ ($1$) to denote a vertex is absent (present) in $I$.
%We use the superscripts to denote which vertex or edge a tensor is associated to.
These labels can be used as subscripts of tensors to index tensor elements, e.g.\ $W^{(v)}_0=1$ is the first element associated with $s_v=0$ and $W^{(v)}_1=x_v^{w_v}$ is the second element associated with $s_v=1$, where $x_v$ is a variable associated with vertex $v$.
$P$ is the output tensor of rank $0$, i.e. a scalar.
The independence constraint is encoded in the edge tensors, where we use $B_{11} = 0$ to denote two vertices connected by an edge cannot be both in the independent set. 
The contraction of this tensor network is
\begin{equation}\label{eq:idp}
    P(G) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{v\in V} W^{(v)}_{s_v} \prod\limits_{(u,v) \in E} B^{(u,v)}_{s_u s_v},
\end{equation}
where the summation runs over all $2^{|V|}$ vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$. 
% It is easy to see that an edge tensor element $B_{s_us_v}$ encodes the independent set constraint. It zeros out the product whenever both $u$ and $v$ are in the independent set.
A vertex tensor element $W^{(v)}_{s_v}$ contributes a multiplicative factor $x_v^{w_v}$ whenever $v$ is in the set.

@@

For a general graph, it is computationally inefficient to evaluate \Cref{eq:idp} by directly summing up the $2^{|V|}$ products.
A better approach to evaluate a tensor network is: to find a good pair-wise tensor contraction order as a binary tree and then contract two tensors at a time by this order.
\begin{theorem}\label{thm:complexreal}
    The tensor network in \Cref{eq:mistensornetwork} for the independent set problem on graph $G = (V,E)$ can be contracted in $\cc = O(|E|)2^{O({\rm tw}(G))}$ number of additions and multiplications.
\end{theorem}
\begin{proof}
Let us denote the line graph, a graph obtained by mapping an edge in the original graph to a vertex and connect two vertices if and only if their associated edges in the original original graph share a common vertex, of the hypergraph representation of a tensor network $\mathcal{N}$ as $L(\mathcal{N})$.
A contraction order of $\mathcal{N}$ corresponds to a tree decomposition of $L(\mathcal{N})$ and the largest intermediate tensor has a rank equal to the width of its tree decomposition~\cite{Markov2008}.
Therefore, an optimal (in terms of space complexity) contraction order corresponds to the tree decomposition of $L(\mathcal{N})$ with the smallest width (or the treewidth ${\rm tw}(L(\mathcal{N})$). The contraction complexity is $O(|\mathcal{N}|)2^{O({\rm tw}(L(\mathcal{N})))}$, where $|\mathcal{N}|$ is the number of tensors in $\mathcal{N}$.
For the independent set problem on graph $G = (V, E)$, the line graph of the hypergraph representation of the tensor network in \Cref{eq:mistensornetwork} is the same as the graph $G$ up to some isolated vertices, hence the contraction complexity of this tensor network with an optimal contraction order is $O(|E|)2^{O({\rm tw}(G))}$.
\end{proof}

@@

\begin{lemma}
    The number of the largest $K$ independent sets of a graph $G = (V, E)$ can be computed in time $O(K\log K)\cc$.
\end{lemma}
The proof is similar to proving \Cref{thm:complexpoly}, except only $k$ largest coefficients are involved in the addition and multiplication.

@@

The configuration enumeration of independent sets include, for example, the enumeration of all independent sets, the enumeration of all MISs, and the enumeration of independent sets with the largest several sizes.
Recall that in the definition of a vertex tensor in \Cref{eq:vertextensor}, variables carry labels, so that one can read out all independent sets directly from the output polynomials.
The multiplication between labeled variables is commutative while the summation of labeled variables forms a set.
Intuitively, one can use a bit string as the representation of a labeled variable and use the bit-wise or $\lor^\circ$ as the multiplication operation.
For example, in a 5-vertex graph, $x_2$ and $x_5$ can be represented as $01000$ and $00001$ respectively and their multiplication $x_2x_5$ can be represented as $01000 \lor^\circ 00001 = 01001$.
To enumerate all independent sets, we designed an algebra on sets of bit strings:
\begin{equation}
\eqname{SN}
\begin{split}
    s \oplus t &= s \cup t\\
    s \odot t &= \{\sigma \lor^\circ \tau \, \mid \, \sigma \in s, \tau \in t\}\\
    \mymathbb{0} &= \{\}\\
    \mymathbb{1} &= \{0^{\otimes |V|}\},
\end{split}\label{eq:set}
\end{equation}
where $s$ and $t$ are each a set of $|V|$-bit strings.
\begin{example}\label{eg:setalgebra}
    For elements being bit strings of length $5$, we have the following set algebra
\begin{equation*}
\begin{split}
    &\{00001\} \oplus \{01110, 01000\} = \{01110, 01000\} \oplus \{00001\} = \{00001,01110, 01000\}\\
    &\{00001\} \oplus \{\} = \{00001\}\\
&\\
    &\{00001\} \odot \{01110, 01000\} = \{01110, 01000\} \odot \{00001\} = \{01111, 01001\}\\
    &\{00001\} \odot \{\} = \{\}\\
    &\{00001\} \odot \{00000\} = \{00001\}.
\end{split}
\end{equation*}
\end{example}

\begin{theorem}\label{thm:complexset}
    The independent sets of graph $G = (V, E)$ be a graph can be enumerated in time $O(|I||V|)\cc$, where $|I|$ is the number of independent sets.
\end{theorem}
\begin{proof}
To enumerate over $I$,  we initialize variable $x_{i}$ in the vertex tensor to $x_i = \{\boldsymbol{e}_{i}\}$, where $\boldsymbol{e}_i$ is a basis bit string of size $|V|$ that has only one non-zero value at location $i$.
The vertex and edge tensors are thus
\begin{equation}
    \left(W^{\rm SN}\right)^{(v)} = \left(\begin{matrix}
        \mymathbb{1} \\
        \{\boldsymbol{e}_{i}\}
    \end{matrix}\right),   
    \qquad \qquad
        \left(B^{\rm SN}\right)^{(u, v)} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}
The contraction of this tensor network is a set with its elements being all independent set configurations. The time complexity of element-wise addition and multiplication is upper bounded by $|I||V|$, where $|V|$ comes from the linear cost of representing a $|V|$-vertex configuration. Combining with \Cref{thm:complexreal}, the overall time complexity is $O(|I||V|)\cc$.
In practise, the huge multiplicative factor $|I|$ only appear in the last step of tensor contraction.
This complexity only upper bounds the actually performance of the algorithm.
\end{proof}

@@

For example, to enumerate only the MISs, we can define a combined element type $s_{k}\infty^k$, where the coefficients follow the algebra in \Cref{eq:set} and the exponents follow the max-plus tropical algebra.
The combined operations become: 
\begin{equation}
\eqname{P1+SN}
\begin{split}
    s_x\infty^x \oplus s_y\infty^y &= \begin{cases}
        (s_x \cup s_y)\infty^{\max(x,y)}, & x = y\\
        s_y\infty^{\max(x,y)}, & x < y\\
        s_x\infty^{\max(x,y)}, & x > y
    \end{cases},\\
    s_x\infty^x \odot s_y\infty^y &= \{\sigma \lor^\circ \tau \mid \sigma \in s_x, \tau \in s_y\}\infty^{x+y},\\
    \mymathbb{0} &= \{\}\infty^{-\infty},\\
    \mymathbb{1} &= \{0^{\otimes |V|}\}\infty^{0}. \label{eq:countingtropicalset}
\end{split}
\end{equation}

@@

\begin{corollary}
    Let $G = (V, E)$ be a graph, one of its maximum independent set can be computed in time $O(|V|)\cc$.
\end{corollary}
The proof is similar to \Cref{thm:complexmis}, except the set algebra is replace by the one in \Cref{eq:singleconfig}, and the time complexity of element-wise addition and multiplication no longer depends on $|\mathcal{I}_\alpha|$.
The linear dependency of $|V|$ in the time complexity can be remove using the back propagation technique, at the cost of additional space.
\begin{theorem}
    Let $G = (V, E)$ be a graph, one of its maximum independent set can be computed in time $\cc$.
\end{theorem}
\begin{proof}
    Let $\alpha(G) = \max\limits_{s \in I} \sum_{v} s_v$ be the maximum independent set size. The optimal configuration can be obtained by differentiating over vertex tensors as
    \begin{equation}
    \begin{cases}
        1, &\frac{\partial \alpha(G)}{\partial s_v} = 1\\
        0, &\frac{\partial \alpha(G)}{\partial s_v} = 0
    \end{cases}.
    \end{equation}
    It can be numerically computed by differentiating over the process of the tropical tensor network contraction, where the backward rules can be found in \Cref{sec:bounding}. The time overhead of computing single optimal solution is constant comparing to only computing the contraction. Thus the computing time complexity should be the same as in \Cref{thm:complextropical}.
\end{proof}

@@

All the solution space properties and the corresponding algebra on unweighted graphs still hold for integer-weighted graphs, while for general weighted graphs, the independence polynomial is not well defined anymore.
For general weighted graphs, it is more useful to know the $k$ maximum weighted sets and their sizes.
They can be computed by the extended tropical algebra, which is a natural generalization of the max-plus tropical algebra:
\begin{equation}
\eqname{T$k$}
\begin{split}
    s \oplus t &= \texttt{largest}(s \cup t, k)\\
    s \odot t &= \texttt{largest}(\{a+b \mid a \in s, b\in t\}, k)\\
    \mymathbb{0} &= -\infty^{\otimes k}\\
    \mymathbb{1} &= -\infty^{\otimes k-1} \otimes 0
\end{split}\label{eq:ext-tropical}
\end{equation}
where $\texttt{largest}(s, k)$ means truncating the set $s$ by only keeping its $k$ largest values.

@@

\subsection{Matching problem}
A $k$-matching in a graph $G=(V,E)$ is a set of $k$ edges that no two of which have a vertex in common.
We map an edge $(u, v) \in E$ to a degree of freedom $\langle u, v\rangle \in \{0, 1\}$ in a tensor network, where $1$ means an edge is in the set and $0$ otherwise.
The tensor network representation for the matching problem $\mathcal{N}_{match}$ can be defined as
\begin{equation}\label{eq:matchtensornetwork}
\begin{split}
    \Sigma &= \{\langle u,v\rangle \mid (u, v) \in E\},\\
    \mathcal{T} &= \{W^{(v)}_{\langle v, N(v, 1)\rangle \langle v, N(v,2) \rangle \ldots \langle v, N(v, d(v))\rangle} \mid v\in V\} \cup \{B^{(u, v)}_{\langle u,v\rangle} \mid (u, v) \in E\},\\
    O &= P,
\end{split}
\end{equation}
where for each $v\in V$, we define a vertex tensor over its neighborhood $N(v)$ as
\begin{equation}
    W^{(v)}_{\langle v, N(v, 1)\rangle \langle v, N(v, 2) \rangle \ldots \langle v, N(v,d(v))\rangle} = \begin{cases}
        1, & \sum_{i=1}^{d(v)} \langle v, N(v, i) \rangle \leq 1,\\
        0, & \text{otherwise},
    \end{cases}
\end{equation}
and for each bond $(u, v) \in E$, we define a rank one tensor as
\begin{equation}
    B^{(u, v)}_{\langle u, v\rangle} = \begin{cases}
    1, & \langle u, v \rangle = 0 \\
    x^{w_{\langle u,v \rangle}}_{\langle u, v\rangle}, & \langle u, v \rangle = 1
\end{cases}.
\end{equation}
Here, $N(v, k)$ is the $k$th vertex in $N(v)$ and $d(v) = |N(v)|$ is the degree of vertex $v$, a label $\langle v, u \rangle$ is equivalent to $\langle u,v\rangle$.
$W$ tensor specifies the constraint that a vertex cannot be shared by two edges in the edge set, and an edge tensor carries the weights.
